Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-19 MANUSCRIPT.tex
edits

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands},
  pdfauthor={Caelum Mroczek; Abraham Springer; Benjamin Lucas},
  pdfkeywords={Base Flow, Base-flow Index, Regionalization, Machine
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\title{Regional-scale Base-flow Index of Gauged and Ungauged Basins in
Drylands}

\authors{Caelum Mroczek\affil{1}, Abraham Springer\affil{1}, Benjamin
Lucas\affil{2}}
\affiliation{1}{School of Earth and Sustainability, Northern Arizona
University, Flagstaff, AZ, USA, }\affiliation{2}{Department of
Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ,
USA, }
\correspondingauthor{Caelum Mroczek}{csm428@nau.edu}

\begin{abstract}
LOREM IPSUM \#\#\#\#\#\#
\end{abstract}

\section*{Plain Language Summary}
LOREM IPSUM \#\#\#\#\#\#

@@

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry
sub-humid systems, account for 40\% of the Earth's land surface. These
regions are home to approximately 2 billion people globally and
constitute the largest terrestrial biome (IUCN, 2019). Despite
supporting diverse ecosystems and human populations, dryland regions
face mounting challenges exacerbated by increasing urbanization,
expanding agricultural activities, and climate-induced amplification of
precipitation patterns (Taylor et al., 2013). This water scarcity is
intensifying due to the compounding effects of climate variability and
increased groundwater extraction (Taylor et al., 2013). In drylands,
groundwater serves as a vital resource for sustaining ecosystems and
meeting human needs (Scanlon et al., 2006; Yao et al., 2018).

Base flow is the sustained portion of streamflow in the absence of
runoff that is derived from groundwater discharge (USGS, 2018). Base
flow is critical to maintaining seasonal low-flow regimes, supporting
aquatic ecosystems, and facilitating the transport of nutrients and
chemicals. Base-flow contribution to streamflow can be highly variable
spatially (Beck et al., 2013; Bosch et al., 2017; Singh et al., 2018),
and temporally (Ficklin et al., 2016; Tan et al., 2020). Increasing
groundwater extraction, changes in land cover/land use, and changes in
precipitation patterns due to climate change affect the timing and
volumes of base flow (Tan et al., 2020; Taylor et al., 2013). Effective
management of water quantity and quality depends on understanding
seasonal and interannual base flow patterns and long-term changes in
base flow behavior.

@@

To address the challenge of quantifying base flow in ungauged
catchments, numerous studies have applied regression and machine
learning methods. (Ahiablame et al., 2013) found that using a regression
model to estimate annual base flow of ungauged catchments was reasonably
easy and accurate. Beck et al. (2013) overcame the nonlinearity of basin
characteristics and improved results of multivariate analyses by using
artificial neural networks to estimate BFI globally. Singh et al. (2018)
implemented a random forest algorithm to predict long-term BFI for
ungauged catchments across New Zealand. The application of machine
learning in past studies highlights its versatility and effectiveness in
addressing ecohydrologic dynamics.

This study develops a technique for estimating BFI in ungauged basins in
Arizona and determines the state's long-term mean BFI. Using a machine
learning model trained with the hydrogeologic characteristics of surface
water basins, we estimate mean BFI in ungauged basins over a period of
record from 1991-2020. Regional patterns and trends in base flow and BFI
of instrumented sites are subsequently investigated. These results
provide novel analyses of low flow processes in Arizona, furthering our
understanding of climate controls on consistent flows in dryland basins.

@@

The state of Arizona, located in the southwest United States, covers a
total area of 295,253 km\(^2\). Arizona is divided into two primary
physiographic provinces: the Colorado Plateau in the northeast, and the
Basin-and-Range region in the west and south. The Central Highlands is a
transition zone consisting of scattered basins separated by the
mountainous foothills of the Mogollon Rim. The Colorado Plateau is
dominated by high-elevation desert with an average elevation of 1,936
masl (6,352 ft). Mean temperatures range from -6\(^\circ\)C
(20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
(23 in) of precipitation. The Basin-and-Range region has a semi-arid to
arid climate with an average elevation of 490 masl (1600 ft). Average
temperature ranges from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
(110\(^\circ\)F) and the region averages 200 mm (8 in) of precipitation
annually. \#\#\#CITATION\#\#\#

@@

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. O'Donnell et al. (2016) found
that 8-10 years of calibration data were necessary to account for
climate variability in paired watershed studies in the region. Following
this criteria, a minimum of 10 years of record was chosen. Any years
missing more than 30 days of streamflow were removed from the possible
record period. This analysis focuses on natural, streamflow-influencing
recharge, excluding streamgages affected by regulation or diversions.
These gages were identified through annual reports on water data
published by the USGS (USGS, 2010). Finally, streamgages on the Colorado
River were excluded from the dataset on the basis that they represent
managed flows shared by Colorado River Compact states. Following these
parameters, 205 USGS streamgages with acceptable periods of record were
used in this study (Figure~\ref{fig-study-area}). Periods of record
ranged from 10 to 112 years, with a median of 28 years.

@@

Annual precipitation and temperature data came from the PRISM climate
group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
(Buban et al., 2020)\textbf{.} Instead of the water year, PRISM data
uses a calendar-year format, which was adopted for consistency in the
water balance. Although this may introduce challenges in the annual
estimates due to inter-annual snow storage, the use of long-term annual
averages is likely to reduce any potential errors (Reitz et al., 2017).

Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid
climatalogical data set (Abatzoglou et al., 2018). TerraClimate uses a
Penman-Monteith approach to generate a reference potential
evapotranspiration (PET). The PET values were calculated assuming a
reference grass surface across the landscape with unlimited water. In
the drylands of the southwestern US, PET typically exceeds precipitation
annually (Zomer et al., 2022).

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used
to derive key basin characteristics: basin area, average slope, and the
proportion of each basin oriented toward north or south aspects. Various
geospatial variables, such as aspect, were disaggregated then averaged
to assess the areal percentage of each sub-variable within individual
HUC 8 basins. By calculating these percentages, we aimed to get a more
comprehensive understanding of landscape composition across space. Land
cover from USGS-NLCD (National Land Cover Database), hydrologic soil
group from SSURGO (Soil Survey Geographic Database), and underlying
geology and karst from USGS were all similarly averaged across the
basins. Aggregating variables to align with the HUC 8 boundaries allowed
for more precise predictions of BFI by integrating spatial variations
within each basin.

@@

Directly measuring base flow and BFI presents unique challenges
(Eckhardt, 2008). The technique chosen to separate base flow has been
shown to affect results, and the choice of base-flow separation method
is subjective since `true' BFI values are not known (Beck et al., 2013).
However, many methods have been developed to estimate these values.
These methods include the use of tracers (Gonzales et al., 2009),
graphical interpolation (Institute of Hydrology, 1980; Sloto et al.,
1996), and digital filters (Arnold et al., 1995; Eckhardt, 2005; Lyne et
al., 1979; Nathan et al., 1990). These techniques have varying levels of
applicability depending on the spatial scale, time span, and the scope
of the study. Comparisons of various base-flow separation techniques
have been made in (Eckhardt, 2005, 2008; Nathan et al., 1990) , but this
study does not explore the superiority of different methods.

Base flow was calculated using a single-parameter, recursive digital
filter technique from (Nathan et al., 1990). This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) (Lyne et al., 1979). Eckhardt (2023) noted that
recursive digital filters lack a physical basis, but as the method is
easy to automate, objective, and repeatable, it is appropriate for a
regional-scale study. The Lyne-Hollick filter has been used in (Arnold
et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008; Singh et
al., 2018). It takes the form of

@@

The implementation of machine learning models to predict hydrologic
indices has been successful in past studies (Rozos et al., 2021; Schmidt
et al., 2020; Singh et al., 2018). In this work, we used the eXtreme
Gradient Boosting (XGBoost) algorithm (Chen et al., 2016) to predict BFI
at ungauged locations using catchment characteristics as predictors. The
XGBoost algorithm is a decision tree-based ensemble algorithm, which can
be adapted for either regression or classification problems. It has
grown in popularity in recent years due to its rapid, efficient, and
scalable nature, and has proven effective in similar environmental
modeling tasks such as streamflow forecasting Ni et al. (2020) and
classifying land use/land cover data (Georganos et al., 2018).

XGBoost uses gradient boosting on decision tree algorithms to leverage
the combination of multiple very low variance models into one overall
prediction. Gradient boosting works sequentially, with the first tree
being learned on the known target values, and each subsequent tree being
learned on the errors of the tree prior. The algorithm then applies a
weighting to each tree based on this boosting to determine how much the
tree will contribute towards the final output. The final prediction is
obtained by the input data traversing all \(n\) weighted trees in the
learned ensemble. The trained XGBoost model is used to predict BFI in
ungauged catchments based on geospatial and hydroclimate predictor
variables (Table~\ref{tbl-predictors}). Certain features were subdivided
according to their areal coverage of each basin (e.g.~land cover was
divided into 16 subdivisions).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2877}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}@{}}
\caption{Basin-characteristic variables used as initial features in
XGBoost model. Starred features are maintained in the final,
dimensionality-reduced model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation* & PRISM & Basin average \\
& Mean Temperature* & PRISM & Basin average \\
& Mean Evapotranspiration* & TerraClimate & Basin average \\
Geospatial & Elevation* & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover* & NLCD & Percent areal coverage \\
& Hydrologic Soil Group* & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

@@

Machine learning models can face issues with overfitting when provided
with a large set of predicting features. This can cause reductions in
performance on unseen data and requires greater computational power and
memory storage (Li et al., 2017). Dimensionality reduction is a powerful
tool in addressing these issues which can be divided into two broad
categories: feature extraction and feature selection. Feature extraction
involves projecting the higher dimensional dataset to a feature space
with less dimensions, though this entails creating a new set of features
that lose the physical meaning of the original feature space. On the
other hand, feature selection maintains a subset of the original
features, all of which retain their physical meaning providing better
readability and interpretability. Here, a supervised features selection
was used to reduce the number of predictors and, as a result, increased
learning performance, decreased computational costs, and reduced
overfitting.

An initial model was trained on the full feature set of 45 predictors
(Table~\ref{tbl-predictors}). We then used a feature importance method
of feature selection to remove noise and less-important features.
Feature importance scores are an approach where importance scores are
assigned to individual features indicating how much the feature
contributed (positively or negatively) to the model (Murdoch et al.,
2019). Here, we use SHAP (SHapley Additive exPlanations) values as the
method of assigning importance values to each feature (Lundberg \& Lee,
2017). The top-10 most-important features
(\textbf{?@suppfig-shap\_values}) were subset and used in training a
subsequent model which saw performance increases and reduced computing
time. This subsequent model was used as the final model in this
analysis.

@@

Statisical analyses were performed on annual BFI and base flow values of
instrumented streamgages. We examined temporal trends using the
Mann-Kendall nonparametric trend test to determine the presence of
monotonic trends (Kendall, 1970; Mann, 1945). The Mann-Kendall test was
used to determine trends in BFI and trend significance of gauged
reaches. This test is widely used in studies of this kind (Ayers et al.,
2019; Ficklin et al., 2016; Woodhouse et al., 2022).

Trends in BFI and base flow were produced using a Mann-Kendall test for
monotonic trends. This test expects that the data is non-parametric and
that the observations are not autocorrelated over time. Observations at
4 streamgages show significant autocorrelation, though only one
streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a
significant trend in BFI. Due to this autocorrelation, the streamgage
was removed from the trend analysis as the autocorrelation may inflate
the variance of the Mann-Kendall statistic leading to an
over/underestimate of the trend (Hamed \& Rao, 1998).

Trends with a \(\rho \le 0.05\) are considered significant.

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records},
  pdfauthor={Caelum Mroczek; Abraham Springer; Neha Gupta; Benjamin Lucas},
  pdfkeywords={Base Flow, Base-flow Index, Regionalization, Machine
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\title{Regional Base-Flow Index in Arid Landscapes Using Machine
Learning and Instrumented Records}

\authors{Caelum Mroczek\affil{1}, Abraham Springer\affil{1}, Neha
Gupta\affil{2}, Benjamin Lucas\affil{3}}
\affiliation{1}{School of Earth and Sustainability, Northern Arizona
University, Flagstaff, AZ, USA, }\affiliation{2}{Department of Hydrology
and Atmospheric Sciences, University of Arizona, Tucson, AZ,
USA, }\affiliation{3}{Department of Mathematics and Statistics, Northern
Arizona University, Flagstaff, AZ, USA, }
\correspondingauthor{Caelum Mroczek}{csm428@nau.edu}

\begin{abstract}
Base flow, sustained by groundwater discharge, is a vital component of
river ecosystems, particularly in drylands, where water resources are
limited. This study analyzes the instrumented streamflow record in
Arizona to assess long-term base-flow index (BFI) trends across gauged
catchments. Results indicate that approximately 32\% of Arizona's
streamflow originates from groundwater discharge, with significant
spatial variability driven by landscape and climatic factors. Base flow
relationships are analyzed with coincident trends in climate variables
such as precipitation, evapotranspiration, and temperature. Spatial and
climatic trends reveal variability in base-flow contributions, providing
insight into groundwater-surface water interactions in arid and
semi-arid landscapes. Building on this analysis, we applied machine
learning methods to predict BFI in ungauged basins, addressing the
challenges of Arizona's sparse streamgage network. Using the eXtreme
Gradient Boosting (XGBoost) algorithm trained on hydroclimate and
physiographic predictors, we estimated long-term BFI from 1991 to 2020.
This combined approach integrates observational data with predictive
modeling to enhance our understanding of base-flow processes and provide
a framework for water resource management in data-limited regions.
\end{abstract}

\section*{Plain Language Summary}
Rivers in drylands, such as Arizona, depend on groundwater contributions
to maintain flow during dry periods. This portion of streamflow, called
base flow, plays a critical role in supporting ecosystems and water
availability in these regions. This study examines long-term patterns of
base flow in Arizona using data from streamflow monitoring stations. We
found that about 32\% of Arizona's river flow comes from groundwater,
though this varies significantly across the state depending on local
climate and landscape features. We also analyzed how base flow is
influenced by changes in precipitation, evapotranspiration, and
temperature, providing insights into how groundwater and surface water
interact in these complex environments. To estimate base flow in basins
without monitoring stations, we used machine learning techniques. By
training a model on data from monitored sites and corresponding
hydroclimate data, we predicted long-term base flow for ungauged areas
across Arizona. This innovative approach addresses the challenge of
Arizona's sparse monitoring network and provides a valuable tool for
understanding and managing water resources in regions with limited data.

@@

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry
sub-humid systems, account for 40\% of the Earth's land surface. These
regions are home to approximately 2 billion people globally and
constitute the largest terrestrial biome (IUCN, 2019). Despite
supporting diverse ecosystems and human populations, dryland regions
face mounting hydrologic challenges exacerbated by increasing
urbanization, expanding agricultural activities, and climate-induced
amplification of precipitation patterns (Taylor et al., 2013). This
water scarcity is intensifying due to the compounding effects of climate
variability and increased groundwater extraction (Taylor et al., 2013).
In drylands, groundwater serves as a vital resource for sustaining
ecosystems and meeting human needs (Scanlon et al., 2006; Yao et al.,
2018).

Base flow is the sustained portion of streamflow in the absence of
runoff that is derived from groundwater discharge (USGS, 2018). Base
flow is critical to maintaining seasonal low-flow regimes, supporting
aquatic ecosystems, and facilitating the transport of nutrients and
chemicals. Base-flow contribution to streamflow can be highly variable
spatially (Beck et al., 2013; Bosch et al., 2017; Singh et al., 2018),
and temporally (Ficklin et al., 2016; Tan et al., 2020). Increasing
groundwater extraction, changes in land cover/land use, and changes in
precipitation patterns due to climate change affect the timing and
volumes of base flow (Tan et al., 2020; Taylor et al., 2013). Effective
management of water quantity and quality depends on understanding
seasonal and interannual base-flow patterns and long-term changes in
base-flow behavior.

@@

Advancements in machine learning provide tools to predict hydrologic
indices in ungauged basins, addressing the limitations of sparse
streamgage networks. To tackle the challenge of quantifying base flow in
ungauged catchments, numerous studies have applied both regression and
machine learning methods. Ahiablame et al. (2013) found that using a
regression model to estimate annual base flow of ungauged catchments was
reasonably easy and accurate. Beck et al. (2013) overcame the
nonlinearity of basin characteristics and improved results of
multivariate analyses by using artificial neural networks to estimate
BFI globally. Singh et al. (2018) implemented a random forest algorithm
to predict long-term BFI for ungauged catchments across New Zealand.
These applications demonstrate the versatility and effectiveness of
machine learning in capturing complex ecohydrologic dynamics and
improving our understanding of groundwater contributions to streamflow.

This study develops a technique for estimating BFI in ungauged basins
across Arizona and evaluates the state's long-term BFI. Regional trends
in base flow and BFI at instrumented sites are also analyzed, with these
trends being linked to coincident trends in precipitation,
evapotranspiration (ET\textsubscript{O}), and temperature over the same
periods. Using a machine learning model trained on the hydrogeologic
characteristics of surface water basins, we estimate the mean BFI for
ungauged basins from 1991 to 2020. This approach helps address the
spatial gaps in the streamgage network, which is relatively sparse
across the state. The results offer novel insights into low-flow
processes in both gauged and ungauged basins, enhancing our
understanding of climate controls on consistent flows in Arizona. This
study contributes to a more comprehensive view of hydrological dynamics
in the context of arid and semi-arid landscapes.

@@

The state of Arizona, located in the southwest United States, covers a
total area of 295,253 km\(^2\). Arizona is divided into two primary
physiographic provinces: the Colorado Plateau in the northeast, and the
Basin-and-Range region in the west and south. The Central Highlands is a
transition zone consisting of scattered basins separated by the
mountainous foothills of the Mogollon Rim. The Colorado Plateau is
dominated by high-elevation desert with an average elevation of 1,936
masl (6,352 ft). Mean temperatures range from -6\(^\circ\)C
(20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
(23 in) of precipitation. The Basin-and-Range region has a semi-arid to
arid climate with an average elevation of 490 masl (1600 ft). Average
temperature ranges from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
(110\(^\circ\)F) and the region averages 200 mm (8 in) of precipitation
annually (Arizona State Climate Office, 2024).

@@

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. Following the findings of
O'Donnell et al. (2016), which determined that 8--10 years of
calibration data are necessary to account for climate variability in
paired watershed studies in the region, a minimum record length of 10
years was required. Additionally, years with more than 30 missing days
of streamflow data were excluded from the analysis. This study focuses
on natural, streamflow-influencing dynamics, so streamgages affected by
regulation or diversions were excluded. The regulated river streamgages
were identified through annual reports on water data published by the
USGS (USGS, 2010). Furthermore, streamgages along the Colorado River
were omitted because they represent managed flows governed by the
Colorado River Compact. After applying these selection criteria, 205
USGS streamgages with acceptable periods of record were included in the
study (Figure~\ref{fig-study-area}). Periods of record ranged from 10 to
112 years, with a median of 28 years.

While our data selection ensures a robust record for this analysis, it
also highlights gaps in spatial coverage that the machine learning model
can address. Arizona has 184 active USGS streamflow stations (as of
2024) across an area of 295,253 km², whereas Indiana (a humid state in
the U.S.) has 189 active stations within an area of 94,326 km². This
translates to a streamgage density of approximately 2.004 gages per
1,000 km² in Indiana, over three times higher than Arizona's density of
0.623 gages per 1,000 km². This stark difference underscores the
relative sparsity of Arizona's streamgage network, particularly in the
context of its larger geographic area and the unique hydrological
challenges posed by its arid and semi-arid landscapes.

@@

Annual precipitation and temperature data came from the PRISM climate
group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
(Buban et al., 2020). Instead of the water year, PRISM data uses a
calendar-year format, which was adopted for consistency in the water
balance. Although this may introduce challenges in the annual estimates
due to inter-annual snow storage, the use of long-term annual averages
is likely to reduce any potential errors (Reitz et al., 2017).

Annual reference evapotranspiration (ET\textsubscript{O}) data came from
TerraClimate, a 4-km grid climatalogical data set (Abatzoglou et al.,
2018). TerraClimate uses a Penman-Monteith approach to generate a
reference evapotranspiration. The ET\textsubscript{O} values were
calculated assuming a reference grass surface across the landscape with
unlimited water. In the drylands of the southwestern US,
ET\textsubscript{O} typically exceeds precipitation annually (Zomer et
al., 2022).

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used
to derive key basin characteristics: basin area, average slope, and the
proportion of each basin oriented toward north or south aspects. Various
geospatial variables, such as aspect, were disaggregated then averaged
to assess the areal percentage of each sub-variable within individual
HUC 8 basins. By calculating these percentages, we aimed to get a more
comprehensive understanding of landscape composition across space. Land
cover was acquired from USGS-NLCD (National Land Cover Database),
hydrologic soil group from SSURGO (Soil Survey Geographic Database), and
underlying geology and karst from USGS were all similarly averaged
across the basins. Aggregating variables to align with the HUC 8
boundaries allowed for more precise predictions of BFI by integrating
spatial variations within each basin.

@@

Directly measuring base flow and BFI presents unique challenges
(Eckhardt, 2008). The technique chosen to separate base flow has been
shown to affect results, and the choice of base-flow separation method
is subjective since `true' BFI values are not known (Beck et al., 2013).
However, many methods have been developed to estimate these values.
These methods include the use of tracers (Gonzales et al., 2009),
graphical interpolation (Institute of Hydrology, 1980; Sloto et al.,
1996), and digital filters (Arnold et al., 1995; Eckhardt, 2005; Lyne et
al., 1979; Nathan et al., 1990). These techniques have varying levels of
applicability depending on the spatial scale, time span, and the scope
of the study. Comparisons of various base-flow separation techniques
have been made in previous studies (e.g. Eckhardt, 2005, 2008; Nathan et
al., 1990); this study does not explore the superiority of different
methods.

Base flow was calculated using a single-parameter, recursive digital
filter technique from Nathan et al. (1990). This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) (Lyne et al., 1979). Eckhardt (2023) noted that
recursive digital filters lack a physical basis, but as the method is
easy to automate, objective, and repeatable, it is appropriate for a
regional-scale study. The Lyne-Hollick filter has been used in multiple
studies (Arnold et al., 2000; Bloomfield et al., 2009; Santhi et al.,
2008; Singh et al., 2018), and it takes the form of

@@

The implementation of machine learning models to predict hydrologic
indices has been successful in past studies (Rozos et al., 2021; Schmidt
et al., 2020; Singh et al., 2018). In this work, we used the eXtreme
Gradient Boosting (XGBoost) algorithm (Chen et al., 2016) to predict BFI
at ungauged locations using catchment characteristics as predictors
Table~\ref{tbl-predictors}. The XGBoost algorithm is a decision
tree-based ensemble algorithm, which can be adapted for either
regression or classification problems. This algorithm iteratively builds
an ensemble of decision trees, where each tree corrects errors from
previous trees to improve predictions (Chen et al., 2016). Its
efficiency, scalability, and robustness have made it increasingly
popular in recent years, with successful applications in environmental
modeling tasks such as streamflow forecasting (Ni et al., 2020;
Szczepanek, 2022) and land use/land cover classification (Georganos et
al., 2018).

XGBoost operates by leveraging gradient boosting on decision tree
algorithms, combining multiple low-variance models to produce a robust
overall prediction. Gradient boosting works iteratively: the initial
tree is trained on the target values, while subsequent trees are trained
on the residual errors of the preceding tree. Each tree is assigned a
weight based on its contribution to reducing error, and these weights
are used to determine the influence of each tree in the final model. The
ultimate prediction is made by aggregating the outputs of all \(n\)
weighted trees in the ensemble. In this study, the trained XGBoost model
is used to predict BFI in ungauged catchments based on geospatial and
hydroclimate predictor variables (Table~\ref{tbl-predictors}). Certain
features were further subdivided according to their areal coverage
within each basin (e.g.~land cover was divided into 16 subdivisions).
This approach allowed the model to capture finer-scale spatial
variability and improve predictive accuracy.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2603}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}@{}}
\caption{Basin-characteristic variables used as initial features in
XGBoost model. Starred features are maintained in the final,
dimensionality-reduced model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation* & PRISM & Basin average \\
& Mean Temperature* & PRISM & Basin average \\
& Reference Evapotranspiration* & TerraClimate & Basin average \\
Geospatial & Elevation* & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover* & NLCD & Percent areal coverage \\
& Hydrologic Soil Group* & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

Our initial training dataset comprised 7,724 observations across 45
variables. To optimize the model's performance, we first conducted an
exhaustive grid search combined with 5-fold cross-validation to identify
the optimal hyperparameter values. The hyperparameters evaluated
included the learning rate (\(\eta\)), minimum split loss (\(\gamma\)),
maximum tree depth, minimum child weight, and the number of trees. While
not an exhaustive list of all possible XGBoost hyperparameters, the
range of values explored provided sufficient variation to ensure the
selection of a high-performing model.The optimal hyperparameters were
determined to be: 700 trees, a learning rate (\(\eta\)) of 0.05, a
minimum split loss (\(\gamma\)) of 0.075, a maximum tree depth of 7, and
a minimum child weight of 5. Using these values, the XGBoost model was
trained on the dataset with 10-fold cross-validation.

\(K\)-fold cross-validation provides an unbiased estimate of a model's
accuracy on unseen data, while also insuring against overfitting or
underfitting. In this approach, the data is randomly divided into \(k\)
folds of equal size. The model is trained on \(k-1\) folds and tested on
the remaining fold, referred to as the validation set. This process is
repeated \(k\) times, with each fold serving as the validation set
exactly once. Each iteration trains an independent model with the same
hyperparameters but using a different subset of training data. By
averaging the model performance across all \(k\) folds, we achieve a
robust and reliable estimate of predictive accuracy. Root mean squared
error (RMSE) was used as the performance metric for both model
optimization and evaluation. RMSE provided a consistent and
interpretable measure of the model's accuracy throughout the training
and validation process.

\includegraphics[width=62.5in,height=\textheight]{images/k-fold_fig.png}

\caption{\label{fig-k-fold}Schematic of \(k\)-fold cross validation. The
dataset is randomly divided into \(k\) stratified folds. Each fold
serves as the validation set once, while the remaining folds are
combined to create a training set for model development. Performance
metrics for the test set are calculated and recorded, and this process
is repeated for all \(k\) folds.}

@@

Machine learning models are prone to overfitting, especially when
provided with a large set of predictive features (Ying, 2019).
Overfitting can degrade model performance on unseen data and increase
the demand for computational resources and memory storage (Li et al.,
2017). Dimensionality reduction offers a robust solution to these
challenges and generally falls into two broad categories: feature
extraction and feature selection.

Feature extraction involves transforming the original dataset into a
lower-dimensional feature space. However, this process generates new
features that often lack the physical interpretability of the original
variables. In contrast, feature selection identifies a subset of the
original features, preserving their physical meaning while improving
model readability and interpretability (Li et al., 2017). In this study,
supervised feature selection was employed to reduce the number of
predictors, which enhanced learning performance, reduced computational
costs, and mitigated overfitting.

To begin, an initial model was trained using the full feature set of 45
predictors (Table~\ref{tbl-predictors}). A feature selection method
based on feature importance was then applied to identify and remove less
relevant and noisy features. Feature importance scores quantify the
contribution of individual features---either positively or
negatively---to the model's predictions (Murdoch et al., 2019). In this
analysis, SHAP (SHapley Additive exPlanations) values were used to
compute feature importance scores (Lundberg \& Lee, 2017).

SHAP values is a method to explain the prediction of an individual
instance by calculating the contribution of each feature to that
prediction. The method is based on coalition game theory and is
discussed further in Lundberg \& Lee (2017). Here, SHAP values are used
for global interpretation of feature importance and feature effects on
the model. Global feature importance is produced by the absolute Shapley
values of each feature across the dataset, providing a list of features
in order of most to least important. Feature effects provide an
indication of the relationship between the value of a predicting feature
and its impact on the prediction.

The ten most important features Figure~\ref{fig-shap_values} were
selected based on their SHAP values and used to train a subsequent model
Table~\ref{tbl-predictors}. This refined model demonstrated improved
performance and reduced computational time compared to the initial
model. The final model, trained on this optimized feature subset, was
ultimately used for the analysis presented here.

@@

Statistical analyses were conducted on annual BFI and base-flow values
from instrumented streamgages to identify temporal trends using the
Mann-Kendall nonparametric trend test (Kendall, 1970; Mann, 1945). This
test detects monotonic trends in datasets that are non-parametric and
assumes the absence of autocorrelation among observations. This test is
widely used in studies of this nature (Ayers et al., 2019; Ficklin et
al., 2016; Woodhouse et al., 2022).

To check for autocorrelation, we applied the Durbin-Watson test, which
revealed significant autocorrelation at four streamgages on an annual
basis. Of these, only one streamgage (09486500 - Santa Cruz River at
Cortaro, AZ) showed a significant trend in BFI. This streamgage was
excluded from the trend analysis, as autocorrelation could inflate the
variance of the Mann-Kendall statistic, potentially leading to biased
trend estimates (Hamed \& Rao, 1998). Trends with a \(\rho \le 0.05\)
are considered significant.

@@

Predicted values of BFI are plotted against observed values for the
entire period of record of the instrumented dataset in
Figure~\ref{fig-actual_predicted} . The agreement between ``out-of-bag''
predictions (blind cross-validation, treating each site as ungauged) and
observed values is acceptable (R\textsuperscript{2} = 0.764) indicating
that the model performs well across the full dataset. The overall RMSE
is 0.129 and the overall percent bias (pbias) is -5.6\%. Model
performance metrics across various classifications are summarized in
Table~\ref{tbl-performance}. These performance metrics demonstrate that
the regional model performs consistently well across different spatial
and climatic classifications. However, the negative pbias values across
all classifications, along with the overall pbias, indicate a systematic
underprediction of BFI by the model. Categories with relatively lower
R\textsuperscript{2} and Nash-Sutcliffe Efficiency (NSE) values also
exhibit higher biases, reflecting weaker model performance in those
specific contexts.
